Kids PM Summary

### 2019-12-20 dustrak_data_cleaning.Rmd
PM cleaning

---
title: "KidsAIR initial PM data work"
author: "Ethan Walker"
date: "Started 19 Nov 2019, Updated 16 Dec 2019"
output: pdf_document
---

@@

```{r, message=FALSE}
library(tidyverse)
library(readxl)
library(naniar)
library(lubridate)
library(zoo)
```

@@

# Load Dusttrak data and clean up
```{r}
# Load full, raw dataset compiled in chunk above
kids_pm_full <- read_rds("Output/kids_pm_full_raw.rds")

@@

# Save pm dataset as RDS and CSV
write_rds(kids_pm_new, "Output/kids_pm_new.rds")
#write_csv(kids_pm_new, "Output/kids_pm_new.csv")
```

@@

# Function to look at PM data across time
```{r}
# Plot PM by datetime to look at trends, gaps, and peaks
pm_time_plot_function <- function(id, location) {
pm_time_plot <- kids_pm %>% 
  filter(home_winter_id == id & area == location) %>% 
  #filter(pm < 30000) %>% 
  arrange(pm_datetime) %>% 
  mutate(pm_sampling_day = as.character(pm_sampling_day),
         day_of_week = factor(day_of_week,
                              levels = c("Sunday", "Monday", "Tuesday", "Wednesday", 
                                         "Thursday", "Friday", "Saturday"))) %>%
  ggplot() + 
    geom_point(aes(pm_datetime, pm, color = day_of_week), size = 1.5) +
    theme_classic() +
    labs(title = id,
         y = expression(paste("PM"[2.5], " (", mu, g/m^3, ")")),
         x = "Datetime",
         color = "Day of week") +
    theme(axis.title.y = element_text(size = 12,
                                      margin = margin(t = 0, r = 20, b = 0, l = 0)),
          axis.title.x = element_text(size = 12),
          axis.text.x = element_blank(),
          axis.text.y = element_text(size = 12, color = "black"),
          axis.line.x = element_line(colour = "black", size = 1), 
          axis.line.y = element_line(colour = "black", size = 1), 
          axis.ticks = element_blank()) +
    scale_color_manual(values = jvPalette)

@@

# Change function input to look at specific home
# Run all of these lines together
pm_time_plot_function(id = "512", location = "WMT")
head(gaps, 25)
head(pm_desc, 25)
tail(pm_desc, 25)

# filter data for a single home
check_data <- kids_pm %>% 
  filter(area == "AK" & home_winter_id == "10")

@@

>>>>

---
title: "KidsAIR initial PM data work"
author: "Ethan Walker"
date: "Started 19 Nov 2019, Updated 19 Dec 2019"
output: pdf_document
---

@@

```{r, message=FALSE}
library(tidyverse)
library(readxl)
library(naniar)
library(lubridate)
library(zoo)
library(knitr)
jv_palette <- c("#330099","#CC0066","#FF6633", 
                 "#0099CC", "#FF9900","#CC6633",
                  "#FF3366", "#33CC99", "#33999")
```

@@

# Load Dusttrak data and do initial clean-up
```{r}
# Load full, raw dataset compiled in chunk above
kids_pm_full <- read_rds("Output/kids_pm_full_raw.rds")

@@

# Save pm dataset as RDS and CSV
#write_rds(kids_pm_new, "Output/kids_pm_new.rds")
#write_csv(kids_pm_new, "Output/kids_pm_new.csv")
```

@@

# Function to look at PM data across time
```{r}
# Plot PM by datetime to look at trends, gaps, and peaks
pm_time_plot_function <- function(id, location) {
pm_time_plot <- kids_pm %>% 
  filter(home_winter_id == id & area == location) %>% 
  #filter(pm < 30000) %>% 
  arrange(pm_datetime) %>% 
  mutate(pm_sampling_day = as.character(pm_sampling_day),
         day_of_week = factor(day_of_week,
                              levels = c("Sunday", "Monday", "Tuesday", "Wednesday", 
                                         "Thursday", "Friday", "Saturday"))) %>%
  ggplot() + 
    geom_point(aes(pm_datetime, pm, color = day_of_week), size = 1.5) +
    theme_classic() +
    labs(title = id,
         y = expression(paste("PM"[2.5], " (", mu, g/m^3, ")")),
         x = "Datetime",
         color = "Day of week") +
    theme(axis.title.y = element_text(size = 12,
                                      margin = margin(t = 0, r = 20, b = 0, l = 0)),
          axis.title.x = element_text(size = 12),
          axis.text.x = element_blank(),
          axis.text.y = element_text(size = 12, color = "black"),
          axis.line.x = element_line(colour = "black", size = 1), 
          axis.line.y = element_line(colour = "black", size = 1), 
          axis.ticks = element_blank()) +
    scale_color_manual(values = jv_palette)

@@

# Change function input to look at specific home
# Run all of these lines together
pm_time_plot_function(id = "137", location = "NN")
head(gaps, 25)
head(pm_desc, 25)
tail(pm_desc, 25)

# filter data for a single home
check_data <- kids_pm %>% 
  filter(area == "NN" & home_winter_id == "140")

@@

# Fix dusttrak sample start times
```{r}
# Load data
kids_pm <- read_rds("Output/kids_pm_new.rds")

# Checking difference between dustrak start datetime and field data log datetime
kids_pm_new2 <- kids_pm_new %>% 
  #filter(area == "NN") %>% 
  group_by(area, home_winter_id) %>% 
  arrange(pm_datetime) %>% 
  unite(start_datetime, c("startdate", "starttime"), 
        sep = " ", remove = FALSE) %>% 
  mutate(first_datetime = first(pm_datetime),
         first_datetime = ymd_hms(first_datetime),
         start_datetime = ymd_hms(start_datetime),
         time_diff = first_datetime - start_datetime) %>% 
  #distinct(time_diff, .keep_all = TRUE) %>% 
  # make general time changes based on ranges
  mutate(pm_datetime = ymd_hms(pm_datetime),
         pm_datetime_new = if_else(time_diff > 1800 & time_diff < 5400, 
                                   pm_datetime - 3600, pm_datetime),
         pm_datetime_new = if_else(time_diff < -1800 & time_diff > -5400, 
                                   pm_datetime_new + 3600, pm_datetime_new),
         pm_datetime_new = if_else(time_diff > 5400 & time_diff < 9000, 
                                   pm_datetime_new - 7200, pm_datetime_new),
         pm_datetime_new = if_else(time_diff < -5400 & time_diff > -9000, 
                                   pm_datetime_new + 7200, pm_datetime_new)) %>% 
  # make specific time changes
  mutate(pm_datetime_new = if_else(time_diff == 10903, # WMA457
                                   pm_datetime_new - 10800, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == (-82721), # WMA440
                                   pm_datetime_new + 82800, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == 10069, # WMA459
                                   pm_datetime_new - 9900, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == 9817, # WMA314
                                   pm_datetime_new - 3600, pm_datetime_new),
         pm_datetime_new = if_else(time_diff < -31535900, # YKC06-103 
                                   pm_datetime_new + 31539519, pm_datetime_new),
         pm_datetime_new = if_else(time_diff < -31532200, # WMA305
                                   pm_datetime_new - 3600, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == 9120, # WMA228
                                   pm_datetime_new - 3600, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == (-21497), # WMA130
                                   pm_datetime_new + 21497, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == 684, # WMA237
                                   pm_datetime_new + 3600, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == 149793, # WMA238
                                   pm_datetime_new - 149793, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == -46227, # CH320
                                   pm_datetime_new + 3600, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == 13514, # TC360 
                                   pm_datetime_new - 3600, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == -10696, # TC357 
                                   pm_datetime_new + 10800, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == -43427, # TC355 
                                   pm_datetime_new + 43200, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == -11388, # TC353 
                                   pm_datetime_new + 10800, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == -221, # TC351 
                                   pm_datetime_new + 3600, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == 4999, # CH214 
                                   pm_datetime_new + 3600, pm_datetime_new),
         pm_datetime_new = if_else(time_diff == 4061, # WMA203 
                                   pm_datetime_new + 3600, pm_datetime_new)) %>% 
  mutate(first_datetime_new = first(pm_datetime_new),
         first_datetime_new = ymd_hms(first_datetime_new),
         start_datetime = ymd_hms(start_datetime),
         time_diff_new = first_datetime_new - start_datetime) %>%  
  distinct(time_diff, .keep_all = TRUE)

kids_pm_new2 <- kids_pm_new %>%
  ungroup() %>% 
  arrange(area, home_winter_id, pm_datetime_new)

# write_rds(kids_pm_new2, "Output/kids_pm_new.rds")
```

# Further cleaning:
## Remove specific homes from PM dataset
## Remove extremely high values
## Change negative values
```{r}
# Load data
kids_pm <- read_rds("Output/kids_pm_new.rds")

kids_pm_new <- kids_pm %>%
  # filtering out specific homes where PM data is not suitable for analysis
  # removes 119,520 obs
         # YKC06_105: only 12 hours of data
  mutate(do_not_use = if_else(area == "AK" & home_winter_id == "39", 1, 0),
         # YKC06_313: sample hourly instead of every minute
         do_not_use = if_else(area == "AK" & home_winter_id == "89", 1, do_not_use),
         # CH_302: sample hourly instead of every minute
         do_not_use = if_else(area == "NN" & home_winter_id == "127", 1, do_not_use),
         # CH_304: Home was resampled
         do_not_use = if_else(area == "NN" & home_winter_id == "1330", 1, do_not_use),
         # CH_306: sample hourly instead of every minute
         do_not_use = if_else(area == "NN" & home_winter_id == "134", 1, do_not_use),
         # CH_307: only 8 datapoints
         do_not_use = if_else(area == "NN" & home_winter_id == "135", 1, do_not_use),
         # TC_357: Home was resampled
         do_not_use = if_else(area == "NN" & home_winter_id == "1510", 1, do_not_use),
         # CH_322: sample hourly instead of every minute
         do_not_use = if_else(area == "NN" & home_winter_id == "161", 1, do_not_use),
         # CH_325: 6300 obs are negative, down to -12.12. Median is negative.
         do_not_use = if_else(area == "NN" & home_winter_id == "165", 1, do_not_use),
         # CH_400: Home was resampled
         do_not_use = if_else(area == "NN" & home_winter_id == "2150", 1, do_not_use),
         # CH_403: Home was resampled
         do_not_use = if_else(area == "NN" & home_winter_id == "2210", 1, do_not_use),
         # CH_108: Only 300 datapoints and no other files found
         do_not_use = if_else(area == "NN" & home_winter_id == "29", 1, do_not_use),
         # CH_207: Only 400 datapoints; no other files found
         do_not_use = if_else(area == "NN" & home_winter_id == "58", 1, do_not_use),
         # TC_230: Home was resampled
         do_not_use = if_else(area == "NN" & home_winter_id == "620", 1, do_not_use),
         # CH_223: Home was resampled
         do_not_use = if_else(area == "NN" & home_winter_id == "840", 1, do_not_use),
         # TC_242: 1 day worth of data with lots of gaps
         do_not_use = if_else(area == "NN" & home_winter_id == "86", 1, do_not_use),
         # WMA277: Home was resampled
         do_not_use = if_else(area == "WMT" & home_winter_id == "1910", 1, do_not_use),
         # WMA305: Home was resampled
         do_not_use = if_else(area == "WMT" & home_winter_id == "2500", 1, do_not_use),
         # WMA306: Home was resampled
         do_not_use = if_else(area == "WMT" & home_winter_id == "2510", 1, do_not_use),
         # WMA358: Home was resampled
         do_not_use = if_else(area == "WMT" & home_winter_id == "2860", 1, do_not_use),
         # WMA448: Home was resampled
         do_not_use = if_else(area == "WMT" & home_winter_id == "3750", 1, do_not_use),
         # WMA200: 4700 negative values, down to -4.85. Median is negative
         do_not_use = if_else(area == "WMT" & home_winter_id == "152", 1, do_not_use),
         # WMA201: 5000 negative values, down to -10. Median negative. 17 hour gap 
         do_not_use = if_else(area == "WMT" & home_winter_id == "153", 1, do_not_use),
         # WMA603: Home was resampled
         do_not_use = if_else(area == "WMT" & home_winter_id == "5100", 
                              1, do_not_use)) %>% 
  filter(do_not_use != 1) %>% 
  # replace PM > 10,000 with NA (33 obs)
  mutate(pm = if_else(pm > 10000, 999999, pm)) %>% 
  replace_with_na(replace = list(pm = 999999)) %>% 
  # replace PM < 1 (lower DustTrak limit) with 0.5 (lower limit / 2) 
  # 291,490 obs less than 1; 56,684 obs less than 0 
  mutate(pm = if_else(pm < 1, 0.5, pm))

# write_rds(kids_pm_new, "Output/kids_pm_new.rds")
```

# New variables to use for analysis
```{r}
# Load data
kids_pm <- read_rds("Output/kids_pm_new.rds")
cleaning_summary <- read_xlsx("Input/cleaning_summary.xlsx") %>% 
  select(area, home_winter_id, final_cleaning_pm, sensitivity_analysis_pm)

kids_pm_new <- kids_pm %>% 
  mutate(home_winter_id = as.numeric(home_winter_id)) %>% 
  left_join(cleaning_summary, by = c("area", "home_winter_id")) %>% 
  select(-samplingdate, -samplingtime, -pickupdate, -pickuptime,
         -dtmin, -dtmax, -dtaverage, -qaqccomments, -data_comments,
         -pm_sampling_day, -pm_mean_daily) %>% 
  # create sampling days based on datetime differences
  group_by(area, home_winter_id) %>% 
  mutate(pm_datetime_new = if_else(is.na(pm_datetime_new), 
                                   pm_datetime, pm_datetime_new),
         first_datetime_new = first(pm_datetime_new),
         first_datetime_new = ymd_hms(first_datetime_new)) %>% 
  mutate(datetime_diff = pm_datetime_new - first_datetime_new,
         sampling_day = if_else(datetime_diff <= 86400, 1, 0),
         sampling_day = if_else(datetime_diff > 86400 & datetime_diff <= 172800, 
                                2, sampling_day),
         sampling_day = if_else(datetime_diff > 172800 & datetime_diff <= 259200, 
                                3, sampling_day),
         sampling_day = if_else(datetime_diff > 259200 & datetime_diff <= 345600, 
                                4, sampling_day),
         sampling_day = if_else(datetime_diff > 345600 & datetime_diff <= 432000, 
                                5, sampling_day),
         sampling_day = if_else(datetime_diff > 432000 & datetime_diff <= 518400, 
                                6, sampling_day),
         sampling_day = if_else(datetime_diff > 518400 & datetime_diff <= 604800, 
                                7, sampling_day),
         sampling_day = if_else(datetime_diff > 604800 & datetime_diff <= 691200, 
                                8, sampling_day),
         sampling_day = if_else(datetime_diff > 691200 & datetime_diff <= 777600, 
                                9, sampling_day),
         sampling_day = if_else(datetime_diff > 777600 & datetime_diff <= 864000, 
                                10, sampling_day),
         sampling_day = if_else(datetime_diff > 864000, 
                                11, sampling_day)) %>% 
  # new vars for daily means and % expected observations for each day
  group_by(area, home_winter_id, sampling_day) %>% 
  mutate(pm_mean_daily = mean(pm),
         percent_daily_obs_pm = n()/1440) %>% 
  ungroup() %>% 
  arrange(area, home_winter_id, pm_datetime_new) %>% 
  select(-first_datetime, -time_diff, -do_not_use)

test <- kids_pm_new %>% 
  filter(percent_daily_obs_pm >= 0.9) %>% 
  summarise(n()/2241695)
# write_rds(kids_pm_new, "Output/kids_pm_new.rds")
# write_csv(kids_pm_new, "Output/kids_pm_new.csv")
```

### 2019-12-20 pm_summary.Rmd
PM cleaning

---
title: "Kids PM Summary"
author: "Ethan Walker"
date: "Started 22 Nov 2019, Updated 16 Dec 2019"
output: pdf_document
---

@@

```{r, message=FALSE}
library(tidyverse)
library(readr)
library(readxl)
library(knitr)
library(lubridate)
jvPalette <- c("#330099","#CC0066","#FF6633", 
                 "#0099CC", "#FF9900","#CC6633",
                  "#FF3366", "#33CC99", "#33999")
```

@@

# Summary stats
## Updated as data is cleaned
```{r}
# summary by area
summary_area <- kids_pm %>%
  # filtering out specific houses during data cleaning
  ## decide for sure if they should be exluded going forward
  mutate(do_not_use = if_else(area == "AK" & home_winter_id == "39", 1, 0),
         do_not_use = if_else(area == "AK" & home_winter_id == "89", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "127", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "1330", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "134", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "135", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "1510", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "161", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "165", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "2150", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "2210", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "29", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "58", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "620", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "840", 1, do_not_use),
         do_not_use = if_else(area == "NN" & home_winter_id == "86", 1, do_not_use),
         do_not_use = if_else(area == "WMT" & home_winter_id == "1910", 1, do_not_use),
         do_not_use = if_else(area == "WMT" & home_winter_id == "2500", 1, do_not_use),
         do_not_use = if_else(area == "WMT" & home_winter_id == "2510", 1, do_not_use),
         do_not_use = if_else(area == "WMT" & home_winter_id == "2860", 1, do_not_use),
         do_not_use = if_else(area == "WMT" & home_winter_id == "3750", 1, do_not_use),
         do_not_use = if_else(area == "WMT" & home_winter_id == "5100", 
                              1, do_not_use)) %>% 
  filter(do_not_use != 1) %>% 
  # filter out high PM values
  filter(pm < 5000) %>% 
  #filter(actual_vs_expected == 1) %>% 
  #filter(pm <0) %>% 
  group_by(area) %>% 
  # Run summary stats
  summarize("n" = n(),
            "Mean PM" = mean(pm, na.rm = TRUE), 
            "SD PM" = sd(pm, na.rm = TRUE),
            "Min PM" = min(pm, na.rm = TRUE), 
            "Median PM" = median(pm, na.rm = TRUE),
            "Max PM" = max(pm, na.rm = TRUE),
            "Mean Sample Days" = mean(pm_sample_interval, na.rm = TRUE),
            "Min Sample Days" = min(pm_sample_interval, na.rm = TRUE), 
            "Max Sample Days" = max(pm_sample_interval, na.rm = TRUE),
            "Mean Sample Obs" = mean(pm_total_observations, na.rm = TRUE),
            "Min Sample Obs" = min(pm_total_observations, na.rm = TRUE), 
            "Max Sample Obs" = max(pm_total_observations, na.rm = TRUE)) %>% 
  # transpose data for ease of reading
  t()
# create column names and make it look nicer
summary <- as.data.frame(summary_area)
my_names <- summary_area[1, ]
colnames(summary) <- my_names
summary_area <- summary %>% 
  rownames_to_column() %>% 
  rename(Statistic = rowname) %>% 
  filter(Statistic != "area") 
kable(summary_area, digits = 2)
```

# Boxplots for PM by study area
```{r, fig.height=6, fig.width=5}
boxplots_kids_pm <- kids_pm %>% 
  filter(pm < 500) %>% 
  ggplot(aes(area, pm)) +
    geom_boxplot(aes(size = 1), lwd = 1.2, colour = "black", 
               fatten = 1, outlier.size = 1.5, width = 0.35) +
    stat_summary(fun.y=mean, geom="point", shape=17, size=3, color="red") +
    theme_minimal() +
    ylim(-100, 500) +
    #geom_hline(yintercept = 50, color = "red", size = 1.2) +
    labs(y = expression(paste("PM"[2.5], " (", mu, g/m^3, ")"))) +
    #labs(aes(x = "", y = paste("48-hour indoor PM "[2.5], " (", mu, g/m^3, ")"))) +
    theme(axis.text.x = element_text(size = 16, colour = "black"),
          axis.text.y = element_text(size = 16, colour = "black"),
          axis.title.y = element_text(size = 16,
                                      margin = margin(t = 0, r = 20, b = 0, l = 0)),
          axis.title.x = element_blank(),
          axis.line.x = element_line(colour = "black", size = 1.2), 
          axis.line.y = element_line(colour = "black", size = 1.2), 
          axis.ticks = element_blank())
boxplots_kids_pm
```

>>>>

---
title: "Kids PM Summary"
author: "Ethan Walker"
date: "Started 22 Nov 2019, Updated 19 Dec 2019"
output: pdf_document
---

@@

```{r, message=FALSE}
library(tidyverse)
library(readr)
library(readxl)
library(knitr)
library(lubridate)
jv_palette <- c("#330099","#CC0066","#FF6633", 
                 "#0099CC", "#FF9900","#CC6633",
                  "#FF3366", "#33CC99", "#33999")
```

@@

# Summary stats
```{r}
pm_summary <- kids_pm %>% 
  # group by: area, treatment, dtid, day_of_week, Sampling_day
  group_by(day_of_week) %>% 
  summarize("n" = n(),
            "Mean PM" = mean(pm, na.rm = TRUE), 
            "SD PM" = sd(pm, na.rm = TRUE),
            "Min PM" = min(pm, na.rm = TRUE), 
            "Median PM" = median(pm, na.rm = TRUE),
            "Max PM" = max(pm, na.rm = TRUE),
            "Mean Sample Days" = mean(pm_sample_interval, na.rm = TRUE),
            "Mean Sample Obs" = mean(pm_total_observations, na.rm = TRUE),
            "Mean % Daily Obs" = mean(percent_daily_obs_pm, na.rm = TRUE))
```

# Boxplots for PM by study area
```{r, fig.height=6, fig.width=5}
boxplots_kids_pm <- kids_pm %>% 
  #filter(pm < 500) %>% 
  ggplot(aes(area, log(pm))) +
    geom_boxplot(aes(size = 1), lwd = 1.2, colour = "black", 
               fatten = 1, outlier.size = 1.5, width = 0.35) +
    stat_summary(fun.y=mean, geom="point", shape=17, size=3, color="red") +
    theme_minimal() +
    #ylim(0, 500) +
    #geom_hline(yintercept = 50, color = "red", size = 1.2) +
    labs(y = expression(paste("PM"[2.5], " (", mu, g/m^3, ")"))) +
    #labs(aes(x = "", y = paste("48-hour indoor PM "[2.5], " (", mu, g/m^3, ")"))) +
    theme(axis.text.x = element_text(size = 16, colour = "black"),
          axis.text.y = element_text(size = 16, colour = "black"),
          axis.title.y = element_text(size = 16,
                                      margin = margin(t = 0, r = 20, b = 0, l = 0)),
          axis.title.x = element_blank(),
          axis.line.x = element_line(colour = "black", size = 1.2), 
          axis.line.y = element_line(colour = "black", size = 1.2), 
          axis.ticks = element_blank())
boxplots_kids_pm
```
